Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife

### 2023-04-14 analysis/paper/paper.Rmd
writing in progress

Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools designed for this purpose have not been available. Hardy's

@@

- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models). Our R package provides a comparable alternative to Hardy QGIS and Landis plugins, more easily integrated into SpaDES or other R workflows. 
- we selected revelstoke because it is topographically complex, so a good challenge. Hardy result in eastern Canadian shield showed better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for. Stream crossing behaviour in Revelstoke landscape could be improved with more a detailed high resolution cost surface. But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain. None of these simple algorithms will project switchbacks on steep slopes. A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. 

>>>>

Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available. Hardy's Landis-II extension [-@hardy_forest_2019; -@hardy_landis-ii_2023] can be integrated with other components of the Landis-II modelling framework [e.g. @labadie_global_2023]. Our implementations of simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provides a comparable alternative to Hardy's method that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

We evaluated performance of these algorithms in a topographically complex region of British Columbia where we anticipated projection of forest road network patterns would be challenging, using a very simple cost surface model. Thus, we were not surprised to less accurate projections of road density and presence than Hardy found in less mountainous regions of eastern Canada with a model that more carefully accounted for water crossing costs [-@hardy_landis-ii_2023]. The stream crossing behaviour of road projection algorithms in Revelstoke could be improved with a more detailed high resolution cost surface, but none of these simple algorithms will project switchbacks on steep slopes that are characteristic of forest roads in mountains. A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. The performance of these algorithms varies with terrain; in topographically complex landscape high resolution cost surfaces that account for the costs of stream and river crossings will address some but not all of the challenges of forest road projection in mountain regions.

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results. The lack of consistent accurate information on the location of resource roads across Canada and elsewhere [@poley_identifying_2022](Poley, ) Better harvest & existing road data would help errors in projections from errors in the input data (cite lit in coops paper)

@@

- we selected revelstoke because it is topographically complex, so a good challenge. Hardy result in eastern Canadian shield showed better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for. Stream crossing behaviour in Revelstoke landscape could be improved with more a detailed high resolution cost surface. But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain. None of these simple algorithms will project switchbacks on steep slopes.  
